%%%
%%% Acknowledgement.
%%%

\svnInfo $Id$

\chapter*{\pdfbookmark[0]{Acknowledgements}{chapter*.3}Acknowledgements\revision}
\addtocontents{toc}{\protect \contentsline {chapter}{Acknowledgements\revision}{\thepage }{}}


First and foremost my special thanks go to Reinhard
Pichler\index{Pichler, Reinhard} and Fang Wei\index{Wei, Fang} for
supervising and proof-reading my thesis, especially to Fang, as she
continued to supervise my thesis even after she had left the
department.
%
Then I would like to thank Katrin Seyr\index{Seyr, Katrin}, Toni
Pisjak\index{Pisjak, Toni} and Markus Pichlmair\index{Pichlmair,
Markus} for their great assistance in bringing
\url{http://skyline.dbai.tuwien.ac.at} to live.
%
Furthermore thanks to all people from the Database and Artificial
Intelligence Group of Vienna University of Technology for
supporting me cordially in various issues.
%
The author likes to thank Donald Kossmann\index{Kossmann, Donald} for
providing the source code for the dataset generator.
%
I also like to acknowledge the discussions I had with Park
Godfrey\index{Godfrey, Park} about the question if the English
language is ready for a new verb ``to skyline''.
%
Furthermore I would like to thank Albin Bucek\index{Bucek, Albin} and
Jens Kober\index{Kober, Jens} from Österreichisches Bundesdenkmalamt
(BDA) for providing me seven computers to run the experiments on.
%
My thanks further goes to my friends Emmanouil ``Manos''
Paissios\index{Paissios, Emmanouil ``Manos''} and Chris
Roschger\index{Roschger, Chris}, for all the lively discussions, the
support, and for proof-reading all the material, and to Nadine
Zollpriester\index{Zollpriester, Nadine} for also proof-reading my thesis.
%
Last but not least I am deeply grateful for all the support I got from
my family, from %my beloved girlfriend 
Wiebke, and our son Paul Isidor,
who was born while I was working on this thesis. I love you all.
